---
title: |-
  S&DS 563 / F&ES 758b - Multivariate Statistics Homework #4
  Cluster Analysis
author: "Lanxin Jiang (lj345), Grace Sun (ys544), Chenglin Lu (cl939)"
date: '`r Sys.Date()`'
output:
  pdf_document:
    latex_engine: pdflatex
geometry: margin=0.5in
header-includes:
   - \usepackage{float}
---

1. Evaluate the assumptions implicit to Discriminant Analysis for your data - multivariate normality WITHIN each group (i.e. chi-square quantile plots) and similarity of covariances matrices (look at Box's M or just look at raw standard deviations/covariance matrices). Comment on what you find. Comment on whether you think transformations might help your data to meet the assumptions of DA. If you think they might, make some transformations and find out! You may also want to make a matrix plot (or a pairs plot) to get a sense of what your data looks like two variables at a time (use different symbols for each group).

    ```{R, echo=F, eval=T}
    setwd("~/Documents/S&DS 563/HW3")
    library(knitr)
    library(pander)
    library(MASS)
    library(biotools)
    library(DiscriMiner)
    library(klaR)
    knitr::opts_chunk$set(fig.align="center")
    # https://archive.ics.uci.edu/ml/datasets/chronic_kidney_disease
    ```

    ```{R}
    Sys.setenv(JAVA_HOME="C:\\Program Files\\Java\\jre-9.0.4\\")
    library(rJava)
    library(RWeka)
    CKD <- read.arff("C:/Users/lanxin/Documents/GitHub/Chronic_Kidney_Disease/Chronic_Kidney_Disease/chronic_kidney_disease_full.arff")
    CKD.cc<-CKD[complete.cases(CKD),]

    ```

use binary variables: appet
use continupus variables: age, bmi

```{r}
#get online function
source("http://www.reuningscherer.net/STAT660/R/CSQPlot.r.txt")
#examine multivariate normality within each group
 CSQPlot(CKD.cc[CKD.cc$appet=="poor",c(1,2,13,15,18)],label="poor")
 CSQPlot(CKD.cc[CKD.cc$appet=="good",c(1,2,13,15,18)],label="good")
```
Multinormal assumption is achieved

```{r}
#make scatter plot to look at differences between groups
plot(CKD.cc[,c(1,2,13,15,18)],col=as.factor(CKD.cc$appet))
```

```{r}
#visually compare sample covariance matrices
print("Covariance Matrix for patients with good appetite")
cov(CKD.cc[CKD.cc$appet=="good",c(1,2,13,15,18)])
print("Covariance Matrix for patients with good appetite")
cov(CKD.cc[CKD.cc$appet=="poor",c(1,2,13,15,18)])

```
```{r}
#calculate Box's M statistic
boxM(CKD.cc[,c(1,2,13,15,18)],CKD.cc[,22])
```
covariance is not different between groups
DA assumes
-Data from both groups comes from a multivariate normal distribution (test with Q-Q Chi-square)
- Covariance matrices in each group are identical (i.e. same shapes, different locations / centroids)
However, the second assumption is violated. Thus, scaling might be essenctial.
Both assumptions are met


2. Perform stepwise discriminant analysis on your data. Comment on which model seems the best. Use quadratic discriminant analysis if appropriate.
```{r}
#STEPWISE DA
step1=stepclass(appet~age+bp+sod+hemo+rbcc, data=CKD.cc, method="lda", direction='both')
step2=stepclass(appet~age+bp+sod+hemo+rbcc, data=CKD.cc, method="lda", direction='both', criteria="AS")
#Do stepwise quadratic DA
step3=stepclass(appet~age+bp+sod+hemo+rbcc, data=CKD.cc,  method="qda", direction='both', criteria="AS")
```
The above stepwise selection shows that the final model should be "appet~rbcc" 

```{r}
#plot results in space spanned by choosen variables
partimat(appet~age+bp+sod+hemo+rbcc,data=CKD.cc,method="lda")
```





3. Comment on whether there is statistical evidence that the multivariate group means are different.
could use manova to comapare multivarite group means
```{r}
manova_ckd <- manova(cbind(age,bp,sod,hemo,rbcc) ~ appet, data = CKD.cc)
summary.manova(manova_ckd,test="Wilks")
summary.aov(manova_ckd)
```
All the variables are significantly different among different appetite. 

4. How many discriminant functions are significant? What is the relative discriminating power of each function?
```{r}
lda_CKD<- lda(appet~age+bp+sod+hemo+rbcc, data=CKD.cc)
lda_CKD
```

```{r}
#normalized coefficient
sum_coef<-sum(m2$scaling^2)
norm_coef<-m2$scaling/sum_coef
scale_coef<-scale(norm_coef)
```
We have only one significant function.  


5. Use classification, both regular and leave-one-out to evaluate the discriminating ability of your functions.
```{r}
pred <- table(CKD.cc$appet, predict(lda_CKD)$class)
# total percent correct
round(sum(diag(prop.table(pred))),2)
```
91% percentage correct by classification. 
```{r}
#cross validated results
lda_CKDCV<- lda(appet~age+bp+sod+hemo+rbcc, data=CKD.cc,CV=TRUE)
CKD_CV <-table(CKD.cc$appet, lda_CKDCV$class)
# total percent correct
round(sum(diag(prop.table(CKD_CV))),2)
```
only 89% correct by leave-one-out cross validation


6. Provide some evidence as to which of your original variables are the 'best' discriminators amongst your groups (look at standardized discriminant coefficients).

We only have one function. It's hard to compare.
We decided to compare by the correction rate of prediction.
```{r}
lda_CKD3<-lda_CKD2<-lda_CKD
lda_CKD2$scaling<-norm_coef
lda_CKD3$scaling<-scale_coef
p1 <- predict(lda_CKD,newdata =  CKD.cc[,c(1,2,13,15,18)])
t1<-table(p1$class, CKD.cc$appet)
print(paste("The correctness rate for the original variables",round(sum(diag(prop.table(t1))),2)))
p2 <- predict(lda_CKD2,newdata =  CKD.cc[,c(1,2,13,15,18)])
t2<-table(p2$class, CKD.cc$appet)
print(paste("The correctness rate for the normalized coefficients",round(sum(diag(prop.table(t2))),2)))
p3 <- predict(lda_CKD3,newdata =  CKD.cc[,c(1,2,13,15,18)])
t3<-table(p3$class, CKD.cc$appet)
print(paste("The correctness rate for the standardized coefficients",round(sum(diag(prop.table(t3))),2)))
```


7. Make a score plot for the first two or three DA function scores (be sure to use different symbols/colors for each group). Comment on what you see.
```{r}
#SCORE PLOTS for linear DA
ckdlda=lda(CKD.cc[,c(1,2,13,15,18)],grouping=CKD.cc$appet)
#Calculate scores
scores=as.matrix(CKD.cc[,c(1,2,13,15,18)])%*%matrix(ckdlda$scaling,ncol=1)
#only one discriminant function
boxplot(scores~CKD.cc$appet,lwd=3, col='red', horizontal=T, main="CKD Discriminant Scores by appetitie")
```
The discriminant analysis of one significant function discriminates appetite between variables well.

8. Bonus (and optional)- try kernel smoothing or k-nearest neighbors and get the admiration of your professor and TA (and some extra credit)! You'll have to use SAS or R for this.
kernel smoothing by leave-one-out cross validation to choose h
Here I use the Nadaraya-Watson kernel estimator,illustrate by two continous variables: age, sc
$$
\hat r_n(x)=\sum_{i=1}^nl_i(x)Y_i
$$
where the kernel function is Gaussian
$$
l_i(x)=\frac{K(\frac{x-x_i}h)}{\sum _{j=1}^nK(\frac{x-x_j}h)}
$$
The leave-one-out cross-validation score $\hat R(h)$ can be written as 
$$
\hat R(h) = \frac 1 n \sum_{i=1}^n (\frac {Y_i - \hat r_n(x_i)} {1-L_{ii}})^2
$$
where $L_{ii} = l_i (x_i)$ is the $i$th diagonal element of the smoothing matrix $L$.

```{r}
locreg<-function(x,y,bandwidth){
  n<-length(x)
  L1<-h_matrix<-matrix(0,ncol=n,nrow=n)
  for(i in 1:n){
    L1[i,]<-dnorm((x-x[i])/bandwidth)
  }
  for (i in  1:n){
    for(j  in 1:n){
      h_matrix[i,j]<-L1[i,j]/sum(L1[,j])
    }  
  }
r_nx<- h_matrix %*% y
score<-1/n*sum((y-r_nx)^2/(1-diag(h_matrix))^2)
return(list(r_nx,score,h_matrix))
}
```

```{r}
#choose bd score
bd_score<-bd<-c(seq(1,20,0.5))
for (i in 1 : length(bd)){
  bd_score[i]<-locreg(CKD.cc$age,CKD.cc$sc,bd[i])[[2]]
}
plot(bd,bd_score)
bd_test<-as.data.frame(cbind(bd,bd_score))
#the best bandwidth of the kernel to minimize the risk
bd_testorder<-bd_test[with(bd_test,order(bd_score)),]
#best bandwidth=10
optimal_bd<-locreg(CKD.cc$age,CKD.cc$sc,10)

plot(CKD.cc$age,CKD.cc$sc,col="red")
points(CKD.cc$age,optimal_bd[[1]])
```

