---
title: |-
  S&DS 563 / F&ES 758b - Multivariate Statistics Homework #6 Option B Factor Analysis
author: "Lanxin Jiang (lj345), Grace Sun (ys544), Chenglin Lu (cl939)"
date: '`r Sys.Date()`'
output:
  word_document: default
  pdf_document:
    latex_engine: pdflatex
header-includes: \usepackage{float}
geometry: margin=0.5in
---
```{R, echo=F, eval=T}
    #setwd("~/Documents/S&DS 563/HW3")
    library(knitr)
    library(pander)
    library(MASS)
    library(biotools)
    library(DiscriMiner)
    library(klaR)
    knitr::opts_chunk$set(fig.align="center")
```
 
    ```{R}
#Import Dataset
    Sys.setenv(JAVA_HOME="C:\\Program Files\\Java\\jre-9.0.4\\")
    library(rJava)
    library(RWeka)
    CKD <- read.arff("C:/Users/lanxin/Documents/GitHub/Chronic_Kidney_Disease/Chronic_Kidney_Disease/chronic_kidney_disease_full.arff")
    ```
    

####1. Look through indicators (questions). Think about which indicators might be related through latent factors. (nothing to turn in here)
   
```{r}
    CKD.numeric <- CKD[,c(1:2,10:18)]
    # Remove Missing Observations
    CKD.numeric <- CKD.numeric[complete.cases(CKD.numeric),]
```
The dataset obtained from the UCI Machine Learning Repository describes the chronic kidney disease status and blood measurement of patients from Apollo Hospitals in India. There are 400 observations in this dataset. The following variables might be related through latent factors. 
			age		-	age	
			bp		-	blood pressure
			bgr		-	blood glucose random
			bu		-	blood urea
			sc		-	serum creatinine
			sod		-	sodium
			pot		-	potassium
			hemo	-	hemoglobin
			pcv		-	packed cell volume
			wbcc	-	white blood cell count
			rbcc	-	red blood cell count
After removal of missing values, our dataset contain 215 observattions. 

####2. Compute the correlation matrix between all indicators (you may want to do this in batches). Comment on relationships you do/do not observe.

```{r}
library(ggcorrplot)
library(dplyr)
CKD.numeric %>%
      cor(use="pairwise.complete.obs", method = "pearson") %>%
      ggcorrplot(hc.order = TRUE, type = "lower",  lab = TRUE,
        outline.col = "white",
        ggtheme = ggplot2::theme_gray,
        colors = c("#6D9EC1", "white", "#E46726"))

```
From the above correlation matrix plot, the more "orange" indicates that two variables more positively correlated, and the more "blue" indicates more negative correlation. Overall, the correlations are high  among variables which indicates groupings of homogeneous sets. We could observe some negative correlation and some positive ones. Sc, bu, bp, pot, bgr, age, or wbcc is negatively associated with hemo, rbcc or sod, while hemo, rbcc or sod is positively related with rbcc, sod, or hemo. There might be more than one latent factor.

####3. Compute KMO or other measure (i.e. just look at matrix produced above) to comment on suitability of data for factor analysis.

```{r}
library(rela)
fact=paf(as.matrix(CKD.numeric))
#KMO Kaiser-Meyer-Olkin (KMO) measure of adequacy
summary(fact)
```

KMO measure is above 0.8, so factor analysis is meritoriously recommended for this dataset. There are 215 cases for 11 parameters. The sample size is fair for factor analysis. Therefore, the dataset is suitable for factor analysis.

####4. Use Principle Components (or appropriate option in Factor Analysis) to decide on a number of latent factors. You can use Scree Plot, eigenvalue>1, or parallel analysis.
```{r}
    library(factoextra)

    pc1 <- princomp(CKD.numeric, cor=TRUE)
    pc1
    # Screeplot
    fviz_screeplot(pc1, choice="eigenvalue")

    source("http://www.reuningscherer.net/STAT660/R/parallel.r.txt")

    # Parallel analysis plot
    parallelplot(pc1)
```

The first two eigenvalues are larger than 1.  The scree plot and the parallel analysis both show that the first two components should retain. Thus, there should be two latent variables. 

####5. Perform a series of factor analyses using orthogonal models. First, try at least two extraction methods (choose from Principle Components, Principle Axis Factoring, Iterative Principle Components, Maximum Likelihood). Use some method for comparing extraction methods to choose a 'best' method (i.e. RMSR or # residuals greater than .05).
```{r}
#extraction method:Maximum Likelihood
fact1=factanal(CKD.numeric,factors=2,rotation="none")
library(psych)
#extraction method:Principle Axis Factoring
fact2=fa(CKD.numeric,nfactors=2,rotate="none", fm="pa")
#extraction method:iterative PCA
fact3=fa(CKD.numeric,nfactors=2,rotate="none", SMC=FALSE, fm="pa")
#function to get RMSR, or proportion of residuals
comp_fa<-function(method,fact){
  #get reproduced correlation matrix
  repro=fact$loadings%*%t(fact$loadings)
  #residual correlation matrix
  resid=cor(CKD.numeric)-repro
  #get root-mean squared residuals
  len=length(resid[upper.tri(resid)])
  RMSR=sqrt(sum(resid[upper.tri(resid)]^2)/len)
  #get proportion of residuals greater than 0.05 in absolute   value
  prop=sum(rep(1,len)[abs(resid[upper.tri(resid)])>0.05])/len 
  out<-paste("Method:", method, ", RMSR:" ,round(RMSR,3),", proportion of residuals greater than 0.05 in absolute value", round(prop,3))
  print(out)
  return(out)
}

out1<-comp_fa("Maximum Likelihood",fact1)
out2<-comp_fa("Principle Axis Factoring",fact2)
out3<-comp_fa("Iterative PCA",fact3)


```
The above three extration methods without any rotation have pretty similar results. By comparing RMSR, maximum likelihood method is a little bit worse than the other two methods, which might result from the fact that these variables don't follow the multivariate normal distribution. Thus, either principle axis factoring or iterative principle component is a better approach. We will use principle axis factoring for the next question.

####6. Once you've chosen an extraction method, try a varimax and/or a quartimax rotation. Pick one of these rotations and discuss the interpretation of the final factors. Make one or more loading plots as appropriate.

```{r}
#extraction method:Principle Axis Factoring
fact21=fa(CKD.numeric,nfactors=2,rotate="varimax", fm="pa")
fact22=fa(CKD.numeric,nfactors=2,rotate="quartimax", fm="pa")
out1<-comp_fa("Varimax",fact21)
out2<-comp_fa("Quartimax",fact22)
#get loading plot for first two factors
plot(fact21$loadings, pch=18, col='red',main="Principle Axis Factoring with Varimax Rotation")
abline(h=0)
abline(v=0)
text(fact21$loadings, labels=names(CKD.numeric),cex=0.8)
plot(fact22$loadings, pch=18, col='red',main="Principle Axis Factoring with Quartimax Rotation")
abline(h=0)
abline(v=0)
text(fact22$loadings, labels=names(CKD.numeric),cex=0.8)
```

The residuals for the two types of rotation are the same because rotation won't change the the fit between the observed and reproduced correlation matrices. 
The above two loading plots seem to be different. Varimax ratation enables each indicator to have a high loading on one and only one factor, while quartimax in addition let indicators have high loading on one factor. 
The loading plot with varimax rotation shows that bu, sc and pot load heavily on factor one; age, wbcc, and bgr load heavily on factor two. The rest load on both factors, and they load heavily on factor one when the rotation is quartimax.
Thus, it is reasonable to conclude that these measurements of patients are a function of two latent factor. 

